---
title: "Getting Started with netballR"
author: "Steve Lane and Craig Moyle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with netballR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Time-stamp: <2018-04-08 10:08:31 (slane)> -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette provides an overview to get you started with using `netballR`. The package ships with the full 2017 season match and player statistics, while the live Champion Data iStats portal still publishes compatible match JSON for current seasons.

*Current fork note*: the package has been updated for the post-2020 super shot scoring model, and `downloadMatch()` now validates competition, round, and game identifiers before requesting data.

# Sourcing Match Data

Data are sourced from `https://mc.championdata.com/data/` using competition, round, and game identifiers. The bundled examples below use the 2017 Super Netball home-and-away competition ID (`10083`) and the 2017 finals competition ID (`10084`). The full processed 2017 datasets are supplied with this package, while current Super Netball seasons, Australian Diamonds internationals, and other live Australian competitions can be queried with the appropriate competition IDs discovered from the current `netball_aus` iStats portal.

If the live endpoint returns a transient HTTP error, `downloadMatch()` will retry before failing. If the response no longer includes a `matchStats` object, the function stops with an explicit error so schema changes are easier to detect.

## Competition discovery pathways

Use `listCompetitionsNetballAus()` for current / active Australian competitions exposed by the Champion Data `netball_aus` application, including Super Netball and Australian Diamonds internationals when they are present in the live catalogue.

Use `anzc_comp_ids` as a historical lookup for ANZ Championship and NZ National Netball League workflows.

Both discovery pathways lead into the same `downloadFixture()` / `downloadMatch()` workflow once you know the relevant `comp_id`.

```{r competition-discovery, eval=FALSE}
competitions <- listCompetitionsNetballAus()
subset(competitions, grepl("Diamonds", competition_name, ignore.case = TRUE))

anzc_comp_ids
```

Use `listCompetitionsNetballAus()` when you want to discover the broader live catalogue exposed by the Champion Data `netball_aus` application before choosing a `comp_id`.

To download statistics from a single match, you use the `downloadMatch` function. As an example, the following code will download the match from round 5, game 3:

```{r get-data-function,eval=FALSE}
library(dplyr)
library(netballR)
round5_game3 <- downloadMatch("10083", 5, 3)

```

The downloaded object is a list, containing detailed statistics (including period-by-period statistics) for the match and players:

```{r source-data,echo=FALSE,warning=FALSE,message=FALSE}
library(dplyr)
library(netballR)
data(round5_game3)

```

```{r display-data}
class(round5_game3)
names(round5_game3)

```

# Tidying Match and Player Statistics

The full match data can be tidied into match and player statistics, grouped by period. Live tidy outputs include the Champion Data `matchId`, which helps distinguish regular-season and finals matches that reuse round/game numbering.

Tidying match statistics using the `tidyMatch` function:

```{r tidying-match}
tidied_match <- tidyMatch(round5_game3)
tidied_match

```

```{r pre-2020-single-match, eval=FALSE}
# For a single ANZ / NZ match, summarise the result from tidy match stats
legacy_result <- matchPoints_pre_2020(tidied_match)
legacy_result
```

Tidying player statistics using the `tidyPlayers` function:

```{r tidying-players}
tidied_players <- tidyPlayers(round5_game3)
tidied_players

```

# Season Data and Ladders

The package still ships with the full 2017 season match and player statistics in tidied format. These have been obtained using the previously described methods, tidied, and then combined by rows to produce a single data frame:

```{r season-2017}
data(season_2017)
season_2017

```

Using a dataset that contains all matches up to a given round in a season means it is easy to reproduce ladder positions. A `ladders` function is provided that can be used on full season data. For example, here is the ladder as it stood at the end of the 2017 home and away season:

```{r ladders}
ladder <- ladders(season_2017, round_num = 14)
ladder

```

The round number is provided above, as the home and away season contained 14 rounds.

# Legacy scoring helpers

For seasons prior to the super shot era, use the `_pre_2020` helpers. These retain the legacy match-points calculation while still exposing the newer quarter-points summary where it is useful for comparison.

```{r ladders-pre-2020, eval=FALSE}
legacy_ladder <- ladders_pre_2020(season_2017, round_num = 14, old_system = TRUE)
legacy_ladder
```

# Development workflow

This fork is maintained with automated tests and a GitHub Actions `R-CMD-check` workflow. For local work, the repository `Makefile` exposes the same core tasks:

```{r dev-workflow, eval=FALSE}
make test
make build
make check
```
